### 2023-06-21 submission/manuscript.Rmd
Ready to submit

Over the past 10 years, the best method of normalizing the sequencing depth of samples characterized by 16S rRNA gene sequencing has been contentious. An often cited paper by McMurdie and Holmes forcefully argued that rarefying the number of sequence counts was "inadmissible" and should not be employed. However, I identified a number of problems with the design of their simulations and analysis that could compromise their results. In fact, when I reproduced and expanded upon their analysis, it was clear that rarefaction was actually the most robust approach for controlling for uneven sequencing effort across samples. Rarefaction limits the rate of falsely detecting and rejecting differences between treatment groups. Far from being "inadmissible", rarefaction is a valuable tool for analyzing microbiome sequence data.

@@

In 2014 Paul McMurdie and Susan Holmes published "Waste not, want not: why rarefying microbiome data is inadmissible" (WNWN) in *PLOS Computational Biology* [@McMurdie2014]. Their paper has had a significant impact on the approaches that microbiome researchers use to analyze 16S rRNA gene sequence data. According to Google Scholar, this paper has been cited more than 2,400 times as of May 2023. There has been a rebuttal of WNWN that showed how rarefaction is beneficial in some cases [@Weiss2017]. I have received correspondence from researchers over the past 10 years asking how to address critiques from reviewers who criticize my correspondents' analysis for using rarefaction (e.g., see https://twitter.com/inanna_nalytica/status/1264299305067786242). I have also received similar comments from reviewers regarding my own work. For example, I received the critique for a preprint that I posted in 2020 that analyzed the practice of removing rare taxa from microbiome analyses [see comments linked to [@Schloss2023sing]; https://web.archive.org/web/20230308191505/http://dalmug.org/filter-review/]. In the process of responding to these reviewers' comments and preparing a manuscript investigating rarefaction and other approaches to control for uneven sequencing effort, I grew to appreciate that there is significant confusion in the field over what is meant by "rarefying" and "rarefaction".

@@

McMurdie and Holmes analyzed the effect of subsampling and other approaches on clustering accuracy using what they called "Simulation A" in their Figure 2A and elsewhere in WNWN. In Simulation A, they investigated the ability to correctly assign simulated microbiome samples to one of two clusters representing two simulated treatment groups. Thankfully, the R code for Simulation A was provided by the authors in the `simulation-cluster-accuracy/simulation-cluster-accuracy-server.Rmd` R-flavored markdown file that was published as Protocol S1 in WNWN. I will outline the simulation strategy below and will reference line numbers from their R-flavored markdown file with "L" as a prefix.

@@

Non-normalized data were used to calculate Bray-Curtis, Euclidean, Poisson, and Weighted UniFrac distances (L793-798, L801-806). Relative abundance data were used to calculate Bray-Curtis, Unweighted UniFrac , and Weighted UniFrac distances (L760-767). Subsampled data were used as input to calculate distances between samples using Bray-Curtis, Euclidean, Unweighted UniFrac, Weighted UniFrac distances, Poisson distances, and top-Mean Squared Difference (L769-775, L793-798, L801-806). DESeq Variance Stabilization normalized data were used to calculate Bray-Curtis, Euclidean, and Weighted Unifrac distances (L793-798). The Upper-Quartile Log-Fold Change normalized data were only used to calculate top-mean squared difference distances (L801-806). WNWN calculated Bray-Curtis [@Legendre2012], Euclidean [@Legendre2012], Unweighted UniFrac [@Lozupone2005], and Weighted UniFrac distances [@Lozupone2007] using the `Phyloseq` R package [@McMurdie2013]. They calculated Poisson distances using the `PoiClaClu` R package [@Witten2011]. The top-mean squared difference was calculated using the `edgeR` R package [@Robinson2009]. The resulting distance matrices were used to cluster the 80 samples into one of two clusters using partitioning around the medoid (PAM), K-means clustering, and hierarchical clustering (L865-879). Although data for all three methods were presented in Protocol S1, only the PAM data were presented in the main manuscript. The accuracy of the clustering assignments was quantified as the fraction of the 80 samples that were assigned to the correct cluster (L887-908). Since some of the subsampling conditions removed samples below a minimum sequencing depth threshold, those samples were counted as mis-clustered samples yielding minimum accuracies below 50%.

@@

Before assessing the impact of the points I critiqued above, I attempted to replicate the results shown in Figures 4 and 5 of WNWN using the authors' code. I created a Conda environment that used the R version and package versions that were as close as possible to those used in WNWN. It was necessary to patch the WNWN's R-flavored markdown file to render the document to be compatible with the overall workflow of this study. I was able to generate figures similar to those presented as WNWN's Figures 4 and 5; my results are shown in this study as Figures 1 and 2, respectively. The differences in results are likely due to differences in software versions and operating systems. It is also worth noting that the published versions of the two figures differ from those included in Protocol S1 within the rendered HTML file (`simulation-cluster-accuracy/simulation-cluster-accuracy-server.html`) and that the figure numbers are one higher in the paper than those generated by the R-flavored markdown file (i.e., Protocol S1 labels the figures as Figures 3 and 4 corresponding to the published Figures 4 and 5). Regardless of the differences, my results were qualitatively similar to that provided in WNWN's Protocol S1.

@@

Each simulated condition was replicated 5 times in WNWN and the paper reported the mean and standard deviation of the replicate clustering accuracies. The relatively small number of replicates accounts for the jerkiness of the lines in WNWN's Figures 4 and 5 (e.g. the Bray-Curtis distances calculated on the DESeqVS normalized data). A better approach would have been to use 100 replicates as this would reduce the dependency of the results on the random number generator's seed. By increasing the number of replicates it was also possible to compare the probability of falsely and correctly clustering samples from the same and different treatment groups together (see points 10 and 11, below). Because the accuracies were not symmetric around the mean accuracy values the median and 95% confidence intervals or interquartile range should have been reported. To test the effect of increasing the number of replicates, I pulled apart the code in `simulation-cluster-accuracy/simulation-cluster-accuracy-server.Rmd` into individual R and bash scripts that were executed using a Snakemake workflow with the same Conda environment I used above. This was necessary since the number of simulated conditions increased 20-fold with the additional replicates. Such intense data processing was not practical within a single R-flavored markdown document. Again, the observed results were qualitatively similar to those generated using the single R-flavored markdown file (Figures 3 and 4). The increased number of replications resulted in smoother lines and allowed me to present empirical 95% confidence intervals. For all analyses in the remainder of this paper, I used 100 randomized replicates per condition.

@@

Close comparison of WNWN's Figure 4 and my version (Figure 3) revealed several difference between the two plots. First, in the WNWN analysis, the accuracies for the Weighted UniFrac distances at the largest effect size (i.e., 3.5) were 1.00 for median sequencing depths of `r oxford_comma(deseq_wunifrac$n_seqs)`. In my version of the analysis, the values for the same sequencing depths were `r oxford_comma(deseq_wunifrac$accuracy)`, respectively. The 95% confidence interval for these accuracies spanned between `r min(deseq_wunifrac$lci)` and `r max(deseq_wunifrac$uci)`. Second, clustering accuracy for Bray-Curtis and Weighted UniFrac distances calculated using DESeq Variance Stabilization normalized OTU counts were different between the original and my simulation at smaller effect sizes and had wide confidence intervals. Inspection of the DESeq Variance Stabilization normalized OTU counts revealed that the method resulted in negative values. It has been suggested that WNWW turned negative DESeq normalized counts to zero [@Weiss2017]; however, I was unable to find code in `simulation-cluster-accuracy-server.Rmd` that made this transformation. In fact, rendering the R-flavored markdown files in WNWN's Protocol S1 generated warning messages when passing the DESeq normalized counts to the Bray-Curtis calculator, which said, "results may be meaningless because data have negative entries in method 'bray'". Although the Weighted UniFrac calculator function did not generate a similar warning message, negative count values would also result in similarly meaningless distances. Both are due to the fact that the distance calculators sum the counts of each OTU in both samples being compared. In contrast, a Euclidean distance does not use a similar sum, but sums the square of the difference between the OTU abundance in each sample. Even if negative values were converted to zeroes, this would effectively the same as removing rare taxa, which could have a significant impact on the shape of the communities [@Schloss2023sing]. To assess the prevalence of negative counts in the simulated data, I quantified the fraction of negative values in the OTU matrix from each simulation and counted the number of simulations where the normalized OTU table had at least one negative value (Figure 5). In general the fraction of negative OTU counts increased with effect size, but decreased with sequencing effort. The fraction of simulations with at least one negative value increased with effect size and sequencing effort. The high frequency of negative OTU counts resulted in highly variable Bray-Curtis and Weighted UniFrac values. It is likely that because the WNWN analysis only used 5 replicates that the large variation in accuracies at high effect sizes was missed initially. For the rest of this reanalysis study, I will only report results using the DESeq-based Variance Stabilization normalization with the Euclidean distance.

@@

The clustering accuracy measurements reported in the body of WNWN were determined using PAM-based clusters while Protocol S1 also includes K-means and hierarchical clustering. Although the data were not displayed in a manner that lent itself to direct comparison in Protocol S1, close inspection of the rendered figures suggested that PAM may not have been the optimal choice in all situations. Rather, K-means clustering appeared to perform better in many simulations. Because the accuracies were the smallest at lower effect sizes, I focused my comparison at the effect size of 1.15. For each set of 100 replicated simulated datasets, I compared the clustering accuracy across clustering methods to see how often each clustering method resulted in the highest accuracy (Figure 7). Indeed, K-means clustering performed better than the other methods. Among all combinations of normalization methods, distance calculations, and read depths, K-means clustering was at least as good as the other methods in `r best_method$kmeans`% of the randomizations (Figure 7); PAM clustering resulted in clustering accuracies as good or better than the other methods in `r best_method$pam`% of the randomizations; and hierarchical clustering (HClust) was at least as good as the other methods for `r best_method$hclust`% of the randomizations. Finally, I specifically compared the clustering accuracies using rarefaction for each of the distance calculations methods using PAM and K-means clustering. Among the `r sum(rarefaction_count)` combinations of distance calculations and read depths, K-means performed better than PAM in `r rarefaction_count[["TRUE"]]` cases with PAM doing better in the `r rarefaction_count[["FALSE"]]` other case (i.e., calculating distances with `r oxford_comma(unique(fails$distance))` using `r oxford_comma(format(fails$n_seqs, big.mark = ","))` sequences). When using subsampled data, K-means clustering performed better than PAM in each case. Because K-means clustering did so much better than PAM clustering in the simulated conditions, I used K-means clustering for the remainder of this study.

@@

In WNWN, the authors quantified the tradeoff between median sequencing depth, the number of samples removed below the threshold, and clustering accuracy (WNWN's Figure 5, my Figure 4). Although the optimal threshold varied by distance metric, normalization method, and sequencing depth, they removed samples whose number of sequences was less than the 15th percentile (L404-419). They acknowledged that this screening step, which was only used with subsampling, would decrease clustering accuracy putting it at a relative disadvantage to the other methods (page 5, column 1, last paragraph). Therefore, it was not surprising that the peak clustering accuracy for their subsampled data was at 85%. Because the true best result would not known *a priori* in an actual microbiome study, it would be impossible for researchers to conduct a sensitivity analysis comparing the tradeoffs between sequencing depth, sample number, and clustering accuracy to select a sequencing depth for their analysis without the risk of P-hacking. The differences in clustering accuracy between subsampling and rarefaction with PAM and K-means clustering indicated that it was necessary to reassess the tradeoff between the library size minimum quantile and clustering accuracy. When using rarefaction, K-means clustering, and only considering conditions with 2,000 or more sequences, there was not a condition where setting a higher threshold resulted in a better accuracy than using all of the samples (Figure 8). These results showed that for modern sequencing depths, using the full datasets with rarefaction and K-means clustering resulted in accuracies that were better than those observed when removing the smallest 15% of the samples from each simulated dataset. When the WNWN Figure 4 was recast with these approaches, rarefaction performed at least as well as any of the other transformations with each distance calculation (Figure 9).

@@

I and others have observed that not using rarefaction can lead to falsely detecting differences between communities when sequencing effort is confounded with the treatment group [@Weiss2017; @Schloss2023rare]. Previous analyses showed that in these situations rarefaction did the best job of controlling the rates of false detection (i.e., Type I errors) and maintaining the statistical power to detect differences (i.e., 1-rate of Type II errors) of differences between groups of samples. Such situations have been observed when comparing communities at different body parts where one site is more likely to generate contaminating sequence reads from the host [@Morris2013]. To determine whether this result was replicated with the WNWN simulation framework, I created a sequencing depth distribution where sequencing depth was fully confounded with the treatment group using both the GlobalPatterns and Log-scaled sequence distributions. To confound the sequencing depth, sequencing depths from one treatment group were drawn from below the median number of sequences of the sequencing distribution and those for the second treatment group were from above the median. To assess the risk of falsely detecting clusters, I compared the clustering accuracies using an effect size of 1.00 using both the confounded and randomized sequencing distributions (see rows 1 and 2 of Figure 14). The samples should have only been assigned to one cluster; however, each of the clustering methods forced the samples into two clusters. So, when there are two groups of 40 samples that do not differ, the best a method could do would be to correctly assign 41 of the 80 samples for an accuracy of `r 41/80`. The false detection risk did not vary by method when the sequencing depth was randomized across treatment groups. Yet, when the sequencing depth was confounded with treatment group, rarefaction was the most consistent normalization method for controlling the generation of spurious clusters. At larger effect sizes the ability to correctly identify two clusters increased when the sequencing depth was confounded with treatment group (Figure 14). At the effect size of 1.15, the rarefied data generated the highest accuracy clusters regardless of whether the sequencing depths were confounded. Although the level of confounding in this simulation was extreme, it highlights the ability of rarefaction to control the false detection rate and the ability to correctly detect clusters relative to the other normalization methods.

@@

The conclusions from WNWN have had a lasting impact on how researchers analyze microbiome sequence data. As I have demonstrated using WNWN's simulation framework, their claims are not supported. The most important points that lead to the difference in our conclusions include the choice of clustering algorithm and arbitrarily selecting a sequencing effort threshold that was used to remove samples. Furthermore, the decision to evaluate the effectiveness of normalization methods based on clustering samples adds a layer of analysis that has been controversial and not widely used. Ultimately, the authors choice of the word "rarefying" has sewn confusion in the field because it is often used in place of the word "rarefaction". As I have demonstrated there were numerous choices throughout the original study that resulted in rarefying/subsampling looking worse than it truly is. In fact, when the data of the current study are taken as a whole, rarefaction is the preferred approach. Short of obtaining the exact same number of sequences from each sample, rarefaction remains the best approach to control for uneven sequencing effort when analyzing alpha and beta diversity metrics. Although beyond the scope of the current study, the same principles would support using rarefaction to control for uneven sequencing effort in shotgun metagenomic sequencing analyses.

Beyond the discussion of whether rarefaction is appropriate for analyzing microbiome data it is worth commenting on WNWN's advice to use DESeq's Variance Stabilization or edgeR's Upper-Quartile Log-fold Change normalization strategies. These methods have been adopted from gene expression analysis to microbiome analysis. Gene expression analysis implicitly assumes that all samples have the same genes since. While this might work in comparing healthy and diseased tissues from a cohort of patients, it does not generalize to those patients' microbiota. Microbial populations are highly patchy in their distribution. Thus, a zero count for gene expression is more likely to represent a gene below the limit of detection whereas a zero count for a microbiome analysis is more likely to represent the true absence of the OTU. In fact, zeroes are less common in gene expression analyses. But because microbiome studies have so many zeroes, it is necessary to add a pseudo-count to all OTUs for both the edgeR and DESeq-based normalizations (L443 and L487, respectively). In WNWN, a pseudo-count of 1 was used. However, this value is arbitrary and the results can vary based on the choice of pseudo-count and the patchiness of the communities being analyzed. Since WNWN was published, compositional approaches have been proposed to account for uneven sequencing and to provide improved interpretability [@Lin2020; @Martino2019; @Quinn2019; @Palarea2015; @Paulson2013; @Gloor2017]. However, these methods also often require the use of pseudo-counts and are not actually insensitive to uneven sequencing [@Schloss2023rare]. Rarefaction is preferred to these alternatives.

The choice of distance metric is a complicated question and the use of six different metrics in WNWN illustrates the challenges of selecting one. Within the ecology literature, Euclidean distances are widely avoided because joint absence is weighted the same as joint presence of taxa [@Legendre2012]. As discussed in point 11, metrics that are based on community membership (i.e., the presence or absence of taxa) performed worse than those than those that were based on community structure (i.e., their relative abundance). For this reason, the Unweighted UniFrac and other metrics like the Jaccard or Sorenson distance coefficients should likely be avoided. The Poisson distance metric is largely novel to the microbial ecology literature and performed no better than the more traditional metrics. In the current analysis, the phylogenetic Weighted UniFrac distance performed comparably to Bray-Cutis distances for clustering or differentiating between communities with adonis. Regardless of the distance calculation employed, they all performed best when using rarefaction relative to the other normalization methods.

Although the authors claim that "Rarefying counts requires an arbitrary selection of a library size minimum that affects downstream inference" (page 8, column 1, point 3), in actual microbiome studies the selection of a sequencing depth is not as arbitrary as the authors claim. Rather, to avoid "P-hacking", researchers pick a set of criteria where they will include or exclude samples prior to testing their data. Examples of criteria might include the presence of a large gap in the sequencing effort distribution, a desire to include poorly sequenced controls, or the *a priori* stipulation of a minimum sequencing effort. To mitigate concerns of arbitrary or engineered minimum library sizes, researchers should indicate the rationale for the threshold they selected.

>>>>

Over the past 10 years, the best method for normalizing the sequencing depth of samples characterized by 16S rRNA gene sequencing has been contentious. An often cited paper by McMurdie and Holmes forcefully argued that rarefying the number of sequence counts was "inadmissible" and should not be employed. However, I identified a number of problems with the design of their simulations and analysis that could compromise their results. In fact, when I reproduced and expanded upon their analysis, it was clear that rarefaction was actually the most robust approach for controlling for uneven sequencing effort across samples. Rarefaction limits the rate of falsely detecting and rejecting differences between treatment groups. Far from being "inadmissible", rarefaction is a valuable tool for analyzing microbiome sequence data.

@@

In 2014 Paul McMurdie and Susan Holmes published "Waste not, want not: why rarefying microbiome data is inadmissible" (WNWN) in *PLOS Computational Biology* [@McMurdie2014]. Their paper has had a significant impact on the approaches that microbiome researchers use to analyze 16S rRNA gene sequence data. According to Google Scholar, this paper has been cited more than 2,400 times as of June 2023. There has been a rebuttal of WNWN that showed how rarefaction is beneficial in some cases [@Weiss2017]; however, the proponents of WNWN appear to be holding sway over the microbiome community. I have received correspondence from researchers over the past 10 years asking how to address critiques from reviewers who criticize my correspondents' analysis for using rarefaction (e.g., see https://twitter.com/inanna_nalytica/status/1264299305067786242). I have also received similar comments from reviewers regarding my own work. For example, I received the critique for a preprint that I posted in 2020 that analyzed the practice of removing rare taxa from microbiome analyses [see comments linked to [@Schloss2023sing]; https://web.archive.org/web/20230308191505/http://dalmug.org/filter-review/]. In the process of responding to these reviewers' comments and preparing a manuscript investigating rarefaction and other approaches to control for uneven sequencing effort, I grew to appreciate that there is significant confusion in the field over what is meant by "rarefying" and "rarefaction".

@@

McMurdie and Holmes analyzed the effect of subsampling and other approaches on clustering accuracy using what they called "Simulation A" in their Figure 2A and elsewhere in WNWN. In Simulation A, they investigated the ability to correctly assign simulated microbiome samples to one of two clusters representing two simulated treatment groups. Thankfully, the R code for Simulation A was provided by the authors in the `simulation-cluster-accuracy/simulation-cluster-accuracy-server.Rmd` R-flavored markdown (R markdown) file that was published as Protocol S1 in WNWN. I will outline the simulation strategy below and will reference line numbers from their R markdown file with "L" as a prefix.

@@

Non-normalized data were used to calculate Bray-Curtis, Euclidean, Poisson, and Weighted UniFrac distances (L793-798, L801-806). Relative abundance data were used to calculate Bray-Curtis, Unweighted UniFrac , and Weighted UniFrac distances (L760-767). Subsampled data were used as input to calculate distances between samples using Bray-Curtis, Euclidean, Unweighted UniFrac, Weighted UniFrac distances, Poisson distances, and top-Mean Squared Difference (L769-775, L793-798, L801-806). DESeq Variance Stabilization normalized data were used to calculate Bray-Curtis, Euclidean, and Weighted Unifrac distances (L793-798). The Upper-Quartile Log-Fold Change normalized data were only used to calculate top-Mean Squared Difference distances (L801-806). WNWN calculated Bray-Curtis [@Legendre2012], Euclidean [@Legendre2012], Unweighted UniFrac [@Lozupone2005], and Weighted UniFrac distances [@Lozupone2007] using the `Phyloseq` R package [@McMurdie2013]. They calculated Poisson distances using the `PoiClaClu` R package [@Witten2011]. The top-mean squared difference was calculated using the `edgeR` R package [@Robinson2009]. The resulting distance matrices were used to cluster the 80 samples into one of two clusters using partitioning around the medoid (PAM), K-means clustering, and hierarchical clustering (L865-879). Although data for all three methods were presented in Protocol S1, only the PAM data were presented in the main manuscript. The accuracy of the clustering assignments was quantified as the fraction of the 80 samples that were assigned to the correct cluster (L887-908). Since some of the subsampling conditions removed samples below a minimum sequencing depth threshold, those samples were counted as mis-clustered samples yielding minimum accuracies below 50%.

@@

Before assessing the impact of the points I critiqued above, I attempted to replicate the results shown in Figures 4 and 5 of WNWN using the authors' code. I created a Conda environment that used the R version and package versions that were as close as possible to those used in WNWN. It was necessary to patch the WNWN's R markdown file to render the document to be compatible with the overall workflow of this study. I was able to generate figures similar to those presented as WNWN's Figures 4 and 5; my results are shown in this study as Figures 1 and 2, respectively. The differences in results are likely due to differences in software versions and operating systems. It is also worth noting that the published versions of the two figures differ from those included in Protocol S1 within the rendered HTML file (`simulation-cluster-accuracy/simulation-cluster-accuracy-server.html`) and that the figure numbers are one higher in the paper than those generated by the R markdown file (i.e., Protocol S1 labels the figures as Figures 3 and 4 corresponding to the published Figures 4 and 5). Regardless of the differences, my results were qualitatively similar to that provided in WNWN's Protocol S1.

@@

Each simulated condition was replicated 5 times in WNWN and the paper reported the mean and standard deviation of the replicate clustering accuracies. The relatively small number of replicates accounts for the jerkiness of the lines in WNWN's Figures 4 and 5 (e.g. the Bray-Curtis distances calculated on the DESeqVS normalized data). A better approach would have been to use 100 replicates as this would reduce the dependency of the results on the random number generator's seed. By increasing the number of replicates it was also possible to compare the probability of falsely and correctly clustering samples from the same and different treatment groups together (see points 10 and 11, below). Because the accuracies were not symmetric around the mean accuracy values the median and 95% confidence intervals or interquartile range should have been reported. To test the effect of increasing the number of replicates, I pulled apart the code in `simulation-cluster-accuracy/simulation-cluster-accuracy-server.Rmd` into individual R and bash scripts that were executed using a Snakemake workflow with the same Conda environment I used above. This was necessary since the number of simulated conditions increased 20-fold with the additional replicates. Such intense data processing was not practical within a single R markdown document. Again, the observed results were qualitatively similar to those generated using the single R markdown file (Figures 3 and 4). The increased number of replications resulted in smoother lines and allowed me to present empirical 95% confidence intervals. For all analyses in the remainder of this paper, I used 100 randomized replicates per condition.

@@

Close comparison of WNWN's Figure 4 and my version (Figure 3) revealed several difference between the two plots. First, in the WNWN analysis, the accuracies for the Weighted UniFrac distances at the largest effect size (i.e., 3.5) were 1.00 for median sequencing depths of `r oxford_comma(deseq_wunifrac$n_seqs)`. In my version of the analysis, the values for the same sequencing depths were `r oxford_comma(deseq_wunifrac$accuracy)`, respectively. The 95% confidence interval for these accuracies spanned between `r min(deseq_wunifrac$lci)` and `r max(deseq_wunifrac$uci)`. Second, clustering accuracy for Bray-Curtis and Weighted UniFrac distances calculated using DESeq Variance Stabilization normalized OTU counts were different between the original and my simulation at smaller effect sizes and had wide confidence intervals. Inspection of the DESeq Variance Stabilization normalized OTU counts revealed that the method resulted in negative values. It has been suggested that WNWW turned negative DESeq normalized counts to zero [@Weiss2017]; however, I was unable to find code in `simulation-cluster-accuracy-server.Rmd` that made this transformation. In fact, rendering the R markdown files in WNWN's Protocol S1 generated warning messages when passing the DESeq normalized counts to the Bray-Curtis calculator, which said, "results may be meaningless because data have negative entries in method 'bray'". Although the Weighted UniFrac calculator function did not generate a similar warning message, negative count values would also result in similarly meaningless distances. Both are due to the fact that the distance calculators sum the counts of each OTU in both samples being compared. In contrast, a Euclidean distance does not use a similar sum, but sums the square of the difference between the OTU abundance in each sample. Even if negative values were converted to zeroes, this would effectively the same as removing rare taxa, which could have a significant impact on the shape of the communities [@Schloss2023sing]. To assess the prevalence of negative counts in the simulated data, I quantified the fraction of negative values in the OTU matrix from each simulation and counted the number of simulations where the normalized OTU table had at least one negative value (Figure 5). In general the fraction of negative OTU counts increased with effect size, but decreased with sequencing effort. The fraction of simulations with at least one negative value increased with effect size and sequencing effort. The high frequency of negative OTU counts resulted in highly variable Bray-Curtis and Weighted UniFrac values. It is likely that because the WNWN analysis only used 5 replicates that the large variation in accuracies at high effect sizes was missed initially. For the rest of this reanalysis study, I will only report results using the DESeq-based Variance Stabilization normalization with the Euclidean distance.

@@

The clustering accuracy measurements reported in the body of WNWN were determined using PAM-based clusters while Protocol S1 also includes K-means and hierarchical clustering. Although the data were not displayed in a manner that lent itself to direct comparison in Protocol S1, close inspection of the rendered figures suggested that PAM may not have been the optimal choice in all situations. Actually, K-means clustering appeared to perform better in many simulations. Because the accuracies were the smallest at lower effect sizes, I focused my comparison at the effect size of 1.15. For each set of 100 replicated simulated datasets, I compared the clustering accuracy across clustering methods to see how often each clustering method resulted in the highest accuracy (Figure 7). Indeed, K-means clustering performed better than the other methods. Among all combinations of normalization methods, distance calculations, and read depths, K-means clustering was at least as good as the other methods in `r best_method$kmeans`% of the randomizations (Figure 7); PAM clustering resulted in clustering accuracies as good or better than the other methods in `r best_method$pam`% of the randomizations; and hierarchical clustering (HClust) was at least as good as the other methods for `r best_method$hclust`% of the randomizations. Finally, I specifically compared the clustering accuracies using rarefaction for each of the distance calculations methods using PAM and K-means clustering. Among the `r sum(rarefaction_count)` combinations of distance calculations and read depths, K-means performed better than PAM in `r rarefaction_count[["TRUE"]]` cases with PAM doing better in the `r rarefaction_count[["FALSE"]]` other case (i.e., calculating distances with `r oxford_comma(unique(fails$distance))` using `r oxford_comma(format(fails$n_seqs, big.mark = ","))` sequences). When using subsampled data, K-means clustering performed better than PAM in each case. Because K-means clustering did so much better than PAM clustering in the simulated conditions, I used K-means clustering for the remainder of this study.

@@

In WNWN, the authors quantified the tradeoff between median sequencing depth, the number of samples removed below the threshold, and clustering accuracy (WNWN's Figure 5, my Figure 4). Although the optimal threshold varied by distance metric, normalization method, and sequencing depth, they removed samples whose number of sequences was less than the 15th percentile (L404-419). They acknowledged that this screening step, which was only used with subsampling, would decrease clustering accuracy putting it at a relative disadvantage to the other methods (page 5, column 1, last paragraph). Therefore, it was not surprising that the peak clustering accuracy for their subsampled data was at 85%. Because the true best threshold would not known *a priori* in an actual microbiome study, it would be impossible for researchers to conduct a sensitivity analysis comparing the tradeoffs between sequencing depth, sample number, and clustering accuracy to select a sequencing depth for their analysis without the risk of p-hacking. The differences in clustering accuracy between subsampling and rarefaction with PAM and K-means clustering indicated that it was necessary to reassess the tradeoff between the library size minimum quantile and clustering accuracy. When using rarefaction, K-means clustering, and only considering conditions with 2,000 or more sequences, there was not a condition where setting a higher threshold resulted in a better accuracy than using all of the samples (Figure 8). These results showed that for modern sequencing depths, using the full datasets with rarefaction and K-means clustering resulted in accuracies that were better than those observed when removing the smallest 15% of the samples from each simulated dataset. When the WNWN Figure 4 was recast with these approaches, rarefaction performed at least as well as any of the other transformations with each distance calculation (Figure 9).

@@

I and others have observed that not using rarefaction can lead to falsely detecting differences between communities when sequencing effort is confounded with the treatment group [@Weiss2017; @Schloss2023rare]. Previous analyses showed that in these situations rarefaction did the best job of controlling the rates of false detection (i.e., Type I errors) and maintaining the statistical power to detect differences (i.e., 1-rate of Type II errors) of differences between groups of samples. Such situations have been observed when comparing communities at different body parts where one site is more likely to generate contaminating sequence reads from the host [e.g., @Morris2013]. To determine whether this result was replicated with the WNWN simulation framework, I created a sequencing depth distribution where sequencing depth was fully confounded with the treatment group using both the GlobalPatterns and Log-scaled sequence distributions. To confound the sequencing depth, sequencing depths from one treatment group were drawn from below the median number of sequences of the sequencing distribution and those for the second treatment group were from above the median. To assess the risk of falsely detecting clusters, I compared the clustering accuracies using an effect size of 1.00 using both the confounded and randomized sequencing distributions (see rows 1 and 2 of Figure 14). The samples should have only been assigned to one cluster; however, each of the clustering methods forced the samples into two clusters. So, when there are two groups of 40 samples that do not differ, the best a method could do would be to correctly assign 41 of the 80 samples for an accuracy of `r 41/80`. The false detection risk did not vary by method when the sequencing depth was randomized across treatment groups. Yet, when the sequencing depth was confounded with treatment group, rarefaction was the most consistent normalization method for controlling the generation of spurious clusters. At larger effect sizes the ability to correctly identify two clusters increased when the sequencing depth was confounded with treatment group (Figure 14). At the effect size of 1.15, the rarefied data generated the highest accuracy clusters regardless of whether the sequencing depths were confounded. Although the level of confounding in this simulation was extreme, it highlights the ability of rarefaction to control the false detection rate and the ability to correctly detect clusters relative to the other normalization methods.

@@

The conclusions from WNWN have had a lasting impact on how researchers analyze microbiome sequence data. As I have demonstrated using WNWN's simulation framework, their claims are not supported. The most important points that lead to the difference in our conclusions include the choice of clustering algorithm and arbitrarily selecting a sequencing effort threshold that was used to remove samples. Furthermore, the decision to evaluate the effectiveness of normalization methods based on clustering samples adds a layer of analysis that has been controversial and not widely used. Ultimately, the authors choice of the word "rarefying" has sewn confusion in the field because it is often used in place of the word "rarefaction". As I have demonstrated there were numerous choices throughout the original study that resulted in rarefying/subsampling looking worse than it truly is. In fact, when the data of the current study are taken as a whole, rarefaction is the preferred approach. Short of obtaining the exact same number of sequences from each sample, rarefaction remains the best approach to control for uneven sequencing effort when analyzing alpha and beta diversity metrics. Although beyond the scope of the current study, the same principles would also support using rarefaction to control for uneven sequencing effort when calculating alpha and beta diversity metrics in shotgun metagenomic sequencing analyses.

Beyond the discussion of whether rarefaction is appropriate for analyzing microbiome data it is worth commenting on WNWN's advice to use DESeq's Variance Stabilization or edgeR's Upper-Quartile Log-fold Change normalization strategies. These methods have been adopted from gene expression analysis to microbiome analysis. Gene expression analysis implicitly assumes that all samples have the same genes. While this might work in comparing healthy and diseased tissues from a cohort of patients, it does not generalize to those patients' microbiota. Microbial populations are highly patchy in their distribution. Thus, a zero count for gene expression is more likely to represent a gene below the limit of detection whereas a zero count for a microbiome analysis is more likely to represent the true absence of the OTU. In fact, zeroes are less common in gene expression analyses. But because microbiome studies have so many zeroes, it is necessary to add a pseudo-count to all OTUs for both the edgeR and DESeq-based normalizations (L443 and L487, respectively). In WNWN, a pseudo-count of 1 was used. However, this value is arbitrary and the results can vary based on the choice of pseudo-count and the patchiness of the communities being analyzed. Since WNWN was published, compositional approaches have been proposed to account for uneven sequencing and to provide improved interpretability [@Lin2020; @Martino2019; @Quinn2019; @Palarea2015; @Paulson2013; @Gloor2017]. However, these methods also often require the use of pseudo-counts and are not actually insensitive to uneven sequencing [@Schloss2023rare]. Rarefaction is preferred to these alternatives.

The choice of distance metric is a complicated question and the use of six different metrics in WNWN illustrates the challenges of selecting one. Within the ecology literature, Euclidean distances are widely avoided because joint absence is weighted the same as joint presence of taxa [@Legendre2012]. As discussed in point 11, distances that are based on community membership (i.e., the presence or absence of taxa) performed worse than those than those that were based on community structure (i.e., their relative abundance). For this reason, the Unweighted UniFrac and other distances like the Jaccard or Sorenson distance coefficients should likely be avoided. The Poisson distance metric is largely novel to the microbial ecology literature and performed no better than the more traditional distances. In the current analysis, the phylogenetic Weighted UniFrac distance performed comparably to Bray-Cutis distances for clustering or differentiating between communities with `adonis2`. Regardless of the distance calculation employed, they all performed best when using rarefaction relative to the other normalization methods.

Although the authors claim that "Rarefying counts requires an arbitrary selection of a library size minimum that affects downstream inference" (page 8, column 1, point 3), in actual microbiome studies the selection of a sequencing depth is not as arbitrary as the authors claim. Rather, to avoid p-hacking, researchers pick a set of criteria where they will include or exclude samples prior to testing their data. Examples of criteria might include the presence of a large gap in the sequencing effort distribution, a desire to include poorly sequenced controls, or the *a priori* stipulation of a minimum sequencing effort. To mitigate concerns of arbitrary or engineered minimum library sizes, researchers should indicate the rationale for the threshold they selected.

WNWN performed a second set of simulations to address the effect of normalization method on the ability to correctly detect differential abundance of OTUs that were randomly selected to have their relative abundances changed (i.e., "Simulation B"). Re-addressing this set of simulations is beyond the scope of the current analysis and others have already contributed critiques [@Weiss2017]. However, many of the same concerns addressed above would apply. Perhaps most important is the fact that if the relative abundance of several OTUs increase, then the relative abundance of all other OTUs would necessarily decrease because the data are compositional [@Gloor2017]. Thus, one would expect every OTU to be differentially abundant. Because of this, it is not possible to truly modify the abundance of a set of OTUs independent of all other OTUs. This is an important limitation of tests of methods attempting to detect differentially abundant OTUs. Regardless, a form of rarefaction could still be employed for detecting differential abundance. One could subsample the data, perform the statistical test, and identify the differentially abundant OTUs. This process could be repeated and the results synthesized. In my experience the largest differences between subsamples are for low relative abundances OTUs, which are unlikely to be statistically significant.
